Tropical Mathematics for the Lambda Calculus

### 2023-01-16 section4.tex
Commit serale

\subsection{Results on Tropical Laurent Series}
{\color{red}

- recall Trop Laurent Series (TLS) tropical roots

- Example 1

- The morphisms of $\LREL_{!}$ can be seen as generalizations of TLS

- Remark: for $f:\Lawv^{X}\to \Lawv^{Y}$, if $X$ is finite, $f$ is a finite variable TLS; if $f$ is written as a finite $\min$, it is a tropical polynomial.

- Maybe Theorem 7 could even go here?? Saying something like: after the remark above, we can prove a result that confirms what seen in Example 1, for the case when $X$ is finite: $f$ can LOCALLY be seen as a polynomial, the degree of such polynomials potentially diverging when $\epsilon\to 0$. Indeed...

- bisognerebbe anche dire due parole sul differenziale e dell'espansione di Taylor come operazione sulle TLS invece che sulle matrici. 
}

Let us start by recalling Example~\ref{ex:famous_ex}.
By plotting its graph {\color{red}vogliamo plottarlo?}, we see first of all that the function is non-decreasing and concave.
It is easy to see that this is actually always the case:

\begin{proposition}\label{prop:nondecr+conc}
 Any tropical Laurent series $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
\end{proposition}

\subsection{$\Lawv^{X}$ as a normed space.}

{\color{red}Move here also the cone part}

@@

Not only the $f$ of Example~\ref{ex:famous_ex} is continuous, but it is also locally Lipschitz on $\BB R_{>0}$.
Actually, we can prove the following:

\begin{theorem}\label{theorem:fepsilon}
 Let $f:\Lawv\to\Lawv$ a tropical Laurent series with matrix $\widehat f:\N\to\Lawv$.
 For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N$ s.t.:
 \begin{enumerate}
  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv$;
  \item If $f(x_0)=\infty$ for some $x_0<\infty$, then $\C F_\epsilon=\emptyset$;
  \item On all $[\epsilon,\infty]$, $f$ coincides with the tropical \emph{polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\widehat f(n)}$.
 \end{enumerate}
 In particular, $f$ is locally Lipschitz on all $\BB R_{>0}$.
\end{theorem}
\begin{proof}
 We can let $\C F_\epsilon:=\set{n\in\N \mid
 \widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(m)+\epsilon \textit{ for all } m<n}$.
\end{proof}

We believe that a variant of such result holds in $\Lawv^X$ for $X$ finite.
We do not know what happens if $X$ is infinite.

Remark that, in coherence with the previous result, in Example~\ref{ex:famous_ex} $f(x)$ is indeed a $\min$ for all $x>0$.
At $x=0$ we have $f(x=0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$ which is \emph{not} a min.
Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
This phenomenon is reminiscent of [Example 7, PCoh]
%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
, which actually motivated our first investigations.

\subsection{$\Lawv^{X}$ as a metric space.}

The norm $\|-\|_\infty$ naturally induces a metric $\|\B x-\B y\|_{\infty}$ over the spaces $\Lawv^{X}$. 
We now study the Lipschitz properties of tropical Laurent series w.r.t.~these metrics.

A first result is the following:
\begin{proposition}\label{prop:troplinear}
All tropical linear functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
\end{proposition}
\begin{proof}[Proof sketch]
Using the fact that $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
the problem reduces to checking that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
%\begin{proof}
%Using $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
%first observe that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\check f_{a,b}-\B x_{a})-(\inf_{a\in X}\check f_{a,b}-\B y_{a})| \leq
%\sup_{a\in X}|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})|\leq 
% \| \B x-\B y\|_{\infty}$.
%\end{proof}

@@

{\color{red}MOVE:
Let us end this section by mentioning another point of view on tropical Laurent series.
$\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete normed cone.
The normed cone structure induces an order on it, called its \emph{cone order}, by setting:
$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
This order makes it a Scott-continuous dcpo.
Furthermore we have:

\begin{proposition}
  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders on the domain and codomain.
\end{proposition}
}

>>>>

\subsection{Results on Tropical Laurent Series}

Recall the definition of \emph{tropical polynomials} and \emph{tropical Laurent series} (of one variable $x\in\Lawv$) given respectively by Equation~\ref{eq:polytrop} and Equation~\ref{eq:defTLS}.
A $x\in[0,\infty)$ is said to be a (finite) \emph{tropical root} of a tropical Laurent series $f:\Lawv\to\Lawv$ iff $f$ is not differentiable at $x$ (w.r.t.\ the usual topology on $\BB R_{\geq0}$.
This is equivalent to ask that the $\inf$ defining $f$ at $x$ is obtained \emph{at least} twice.{\color{red} \`e vero anche per TLS o solo per poly questo? Inoltre, Robol parla anche di finite end-points of the domain: nel nostro caso sarebbe $x=0$}

\begin{example}\label{ex:famous_ex}
 The function $f:\Lawv\to\Lawv$ defined by $f(x):=\inf\limits_{n\in\N}\set{nx+\frac{1}{2^n}}$ is a tropical Laurent series.
\end{example}

By plotting its graph {\color{red}vogliamo plottarlo?}, we observe several properties that we will lift to the more general case that we will consider in the next lines, and this example will serve as running one.

\begin{remark}%[Tropical Laurent series]
 As usual, a matrix $t\in\HOM{\LREL_!}{X}{Y}$ operates as a linear map $\widehat t:\Lawv^{!X}\to\Lawv^Y$.
 But we can also ``express it in the base $X$'', i.e.\ see it as a map $t^!:\Lawv^X\to\Lawv^Y$, by setting $t^!(x):=t\circ_! x$ (we are identifying $\Lawv^X$ with the set $\HOM{\LREL_!}{\emptyset}{X}$ of the \emph{points} of $X$).
 This is the notion of \emph{non-linear} map generated by the CCC-structure of $\LREL_!$.
 Concretely, we have:
 \[t^!(x)_b=\inf\limits_{\mu\in !X} \set{\mu x+ t_{\mu,b}}\] where $\mu x:=\sum\limits_{a\in X} \mu(a)x_a$.
 The functions of such shape are generalisations of usual tropical Laurent series: they are the case for infinite variables.
 We will call them simply \emph{tropical Laurent series (tLs)}.
 %Since in the general case of $\QREL$, $t^!$ is a Laurent series with operations in $Q$, let us call \emph{tropical Laurent series} the functions of shape $t^!$ for some $t\in\HOM{\LREL_!}{X}{Y}$.
\end{remark}

We find the usual notion of tLs of one variable as follows:

\begin{remark}
 Indentifying $!\set{*}\simeq \N$ and $\Lawv^{\set{*}}\simeq\Lawv$, the tropical Laurent series generated by the morphisms in $\HOM{\LREL_!}{\set{*}}{\set{*}}$ are exactly the functions $f:\Lawv\to\Lawv$ of shape $f(x)=\inf\limits_{n\in\N}\set{nx+\widehat f(n)}$, for some $\widehat f:\N\to\Lawv$, i.e.\ usual tLs's of one variable.
Remark that we also find usual \emph{tropical polynomials} of tropical geometry as a particular case: they corrispond to the ones for which the support $\set{n\in\N\mid\widehat f(n)\neq\infty}$ of $\widehat f$ is \emph{finite}.
Actually, for us \emph{tropical polynomial} will mean a function $f:\Lawv^X\to\Lawv^Y$ of shape $f(x)_b:=\min\limits_{\mu\in\C F} \set{\mu x+ t_{\mu,b}}$, for a \emph{finite} set $\C F\subseteq \, !X$.
This is again the generalisation of usual tropical polynomials to the case of infinitely many variables.
\end{remark}

\begin{remark}
 The $f$ of Example~\ref{ex:famous_ex} is indeed of shape $f=t^!$, for $t\in\Lawv^{!\set{*}\times\set{*}}$, $t_{\mu,*}:=2^{-\# \mu}$.
 By Proposition~\ref{prop:descrete}, $f$ is not the interpretation of a $\lam$-term, because its matrix is not descrete.
 Therefore $\LREL_!$ is not a full-complete model of $\STLC$.
\end{remark}

Looking at Example~\ref{ex:famous_ex}, we see that the function is non-decreasing and concave.
It is easy to see that this is actually always the case:

\begin{proposition}\label{prop:nondecr+conc}
 Any tLs $f:\Lawv^X\to\Lawv^Y$ is non-decreasing and concave, w.r.t.\ the pointwise order.
\end{proposition}

In Example~\ref{ex:famous_ex}, it can be proven by hand that $f(x)$ is a $\min$ for all $x>0$.
At $x=0$ we have $f(x=0)=\inf\limits_{n\in\N} \frac{1}{2^n}=0$ which is thus the only point \emph{not} being a $\min$.
Also, while the derivative of $f$ is bounded on all $\BB R_{>0}$, at $x=0$ it tends to $\infty$.
This phenomenon is reminiscent of [Example 7, PCoh],
%Differentials and Distances in Probabilistic Coherence Spaces. FSCD 2019
which actually motivated our first investigations.
In fact, these properties are shared by all tLs with \emph{finitely} many variables, as shown by the following result.
Remark that $!\set{1,\dots,k}$ can be identified with $\N^k$.
So the matrix of a tLs $f$ with finitely many variables $x=x_1,\dots,x_k$ (and one variable $f(x)$ in output) can be given as a $\widehat f:\N^k\to\Lawv$, and $f$ has shape $f(x)=\inf\limits_{n\in \N^k}\set{nx+\widehat f(n)}$, where $nx$ is the scalar product.

\begin{theorem}\label{theorem:fepsilon}
 Let $k\in\N$ and $f:\Lawv^k\to\Lawv$ a tLs with matrix $\widehat f:\N^k\to\Lawv$.
 For all $0<\epsilon<\infty$, there is a \emph{finite} $\C F_\epsilon \subseteq \N^k$ s.t.:
 \begin{enumerate}
  \item If $\C F_\epsilon=\emptyset$ then $f=\infty$ on all $\Lawv^k$;
  \item If $f(x_0)=\infty$ for some $x_0\in[0,\infty)^k$, then $\C F_\epsilon=\emptyset$;
  \item On all $[\epsilon,\infty]^k$, $f$ coincides with the tropical \emph{polynomial} $P_\epsilon(x):=\min\limits_{n\in \C F_\epsilon}\set{nx+\widehat f(n)}$.
 \end{enumerate}
\end{theorem}
\begin{proof}
 Call $\leq$ the well-founded pointwise order on $\N^k$.
 Now we can let $\C F_\epsilon:=\set{n\in\N \mid
 \widehat f(n)<\infty \textit{ and } \widehat f(m)> \widehat f(n)+\epsilon \textit{ for all } m<n}$.
\end{proof}

{\color{red}
- bisognerebbe anche dire due parole sul differenziale e dell'espansione di Taylor come operazione sulle TLS invece che sulle matrici. 
}

\subsection{$\Lawv^{X}$ as a normed cone}

$\Lawv^X$ with the usual $+$ and the usual $\cdot$ is a $\BB R_{\geq0}$-semimodule.
Together with the norm $\norm{\cdot}_\infty$, it can be proved that it is a Scott-complete normed cone.
The normed cone structure induces an order on it, called its \emph{cone order}, by setting:
$x\leq y$ iff $y=x+z$ for some (unique) $z\in\Lawv^X$.
This order makes it a Scott-continuous dcpo.
Furthermore we have:

\begin{proposition}
  Tropical Laurent series $\Lawv^X\to\Lawv^Y$ are Scott-continuous on $\BB R_{>0}^X$, w.r.t.\ the cone orders on the domain and codomain.
\end{proposition}

@@

\subsection{Metric results}%$\Lawv^{X}$ as a metric space.}

The norm $\norm{\cdot}_\infty$ naturally induces a metric $\|\B x-\B y\|_{\infty}$ over the spaces $\Lawv^{X}$. 
We now study the Lipschitz properties of tropical Laurent series w.r.t.~these metrics.

First of all, remark that Theorem~\ref{theorem:fepsilon} implies that tLs $f$ of finitely many variables are locally Lipschitz (and thus continuous) on all $\BB R_{>0}$.
In this subsection we will refine and generalise this result.

A first result is the following:

\begin{proposition}\label{prop:troplinear}
All tropical linear functions $f: \Lawv^{X}\to \Lawv^{Y}$ are non-expansive.  
\end{proposition}
\begin{proof}[Proof sketch]
Using the fact that $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
the problem reduces to checking that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$.\end{proof}
This result shows that, in analogy with that happens in usual metric semantics, linear programs are interpreted by non-expansive functions. 
%\begin{proof}
%Using $f(\B x)_{b}= \inf_{a\in X}\check f_{a,b}+\B x_{a}$,
%first observe that $|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})| = |\B x_{a}-\B y_{a}|\leq \| \B x-\B y\|_{\infty}$; we now have
%$|f(\B x)_{b}-f(\B y)_{b}| \leq |(\inf_{a\in X}\check f_{a,b}-\B x_{a})-(\inf_{a\in X}\check f_{a,b}-\B y_{a})| \leq
%\sup_{a\in X}|(\check f_{a,b}-\B x_{a})- (\check f_{a,b}-\B y_{a})|\leq 
% \| \B x-\B y\|_{\infty}$.
%\end{proof}

@@
